---
title: "bascule"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bascule}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  include = TRUE
)
```

### Installing the package

The `bascule` R package can be installed directly from GitHub using `devtools`:

```{r install_pkg}
devtools::install_github("caravagnalab/bascule")
```

Once installed, load the package in your R session:

```{r load_pkg, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
library(bascule) 
```

### Installing Python dependencies

The `bascule` package relies on the Python module `pybascule`. The recommended approach is to install these Python dependencies inside a dedicated conda environment.

This can be done from R using the `reticulate` package through the following steps.

- if Miniconda is not already available on your system, you can install it using through [`reticulate`](https://rstudio.github.io/reticulate/reference/install_miniconda.html):

```{r, eval=FALSE}
install_miniconda(path = miniconda_path(), update = TRUE, force = FALSE)
```

- create a new conda environment with Python version 3.10:

```{r}
reticulate::conda_create(envname="bascule", python_version="3.10")
```

- activate the new conda environment within the current R session:

```{r}
reticulate::use_condaenv("bascule")
```

- install the `pybascule` Python module using pip inside the conda environment:

```{r}
reticulate::conda_install(envname="bascule", packages="pybascule", pip=TRUE)
```

## Fitting a `bascule` object

Before fitting a `bascule` object with the `fit()` or the `fit_clustering()` functions, make sure that the correct conda environment is active so that the `pybascule` module can be accessed:

```{r}
reticulate::use_condaenv("bascule")
```

Functions that rely on `pybascule` include an additional argument, `py`, which allows you to explicitly pass the imported Python module. This ensures that the correct version of `pybascule` is used.

```{r, eval=FALSE}
reticulate::use_condaenv("bascule")
py = reticulate::import("pybascule")
x = fit(..., py=py)
# or
x = fit_clustering(..., py=py)
```
