---
title: "survival analysis: skin cancer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{survival analysis: skin cancer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


### Load package

```{r setup}
library(dplyr)
library(survival) # Surv, survfit
library(survminer) # ggsurvplot
library(forestmodel)
#library(bascule)
devtools::load_all("/Users/azadsadr/Documents/packages/bascule/")
```


### Load the BASCULE fit object and clinical metadata

We utilized a fit object of the skin cancer dataset from the International Cancer Genome Consortium (ICGC) cohort, which is readily available within the package and can be loaded using the command `data("skin_fit")`.

Subsequent post-processing involved merging related clusters and mapping the de novo inferred mutational signatures to a predefined set of known signatures from the COSMIC (Catalogue of Somatic Mutations in Cancer) database and Degasperi et al. study.

```{r message=FALSE, warning=FALSE}
#data("synthetic_data")
#load("/Users/azadsadr/Documents/skin_data.RData")
data("skin_fit")
data <- skin_fit #readRDS("/Users/azadsadr/Nextcloud/bascule/revision4/fit_final/02/skin.Rds")
x_orig = merge_clusters(data, cutoff=0.9)
x = convert_dn_names(x_orig, reference_cat=list(SBS=Degasperi_SBS), cutoff=0.8)
x = convert_dn_names(x, reference_cat=list(SBS=COSMIC_sbs_filt), cutoff=0.8)
```

Clinical metadata corresponding to the skin cancer cohort were imported using `data("skin_clinical")` and matched to the genomic data based on unique patient identifiers. The clinical metadata includes the following variables:

* PATIENT: Unique patient identifier (ID)
* TISSUE: Cancerous tissue type
* GENDER: Biological sex of the patient
* OS_STATUS: Overall survival status (alive or deceased)
* OS_YEARS: Time to event (measured as the duration until death, if applicable)
* CLUSTER: Cluster assignment as determined by the BASCULE algorithm
* AGE: Age of the patient at diagnosis

```{r message=FALSE, warning=FALSE}
data("skin_clinical")

sync_metadata <- function(x, metadata) {
  #clinical_data <- readRDS(metadata_path) %>% filter(TISSUE=="Skin")
  clinical_data <- metadata
  exp <- get_exposure(x, add_groups = TRUE)[["SBS"]] %>% 
    group_by(clusters) %>%  # group by cluster
    filter((n_distinct(samples) >= 20), value > 0) %>%   # keep only clusters with 20 or more members
    ungroup()
  
  # Quality check --------------------------------------------------------------
  clusters_list <- exp %>% pull(clusters) %>% unique 
  num_clusters <- clusters_list %>% length
  if (num_clusters == 1) {
    message("There is just one cluster, you may not need to proceed anymore!")
  } else if (num_clusters < 1) {
    stop("Something went wrong: No valid clusters found!")
  }
  
  samples_list <- intersect(
    x = unique(exp$samples), 
    y = clinical_data %>% filter(TISSUE=="Skin") %>% pull(PATIENT) %>% unique
  )
  if (length(samples_list) == 0 ) stop("ERROR: No common samples found between exposure and clinical data!")
  
  clinical_data <- clinical_data %>% filter(PATIENT %in% samples_list)
  exp <- exp %>% filter(samples %in% samples_list)
  
  # Preprocess data ------------------------------------------------------------
  cluster_assignment <- exp %>% select(samples, clusters) %>% distinct # get unique sample-cluster assignment
  df <- merge(x = clinical_data, y = cluster_assignment, by.x="PATIENT", by.y = "samples", all.x = TRUE) # merge with clinical data
  
  # Convert variables ----------------------------------------------------------
  df$GENDER <- as.factor(df$GENDER)
  
  df$OS_STATUS <- ifelse(df$OS_STATUS == "DECEASED", 1, 0)
  
  names(df)[names(df) == "clusters"] <- "CLUSTER"
  df$CLUSTER <- as.factor(df$CLUSTER)
  #if (!is.null(clusters_list)) {
  #  levels(df$CLUSTER) <- clusters_list
  #}
  
  df <- df %>% mutate(AGE_GROUP = ifelse(df$AGE > median(df$AGE), "old", "young"))
  df$AGE_GROUP <- as.factor(df$AGE_GROUP)
  
  df <- df %>% filter(!is.na(CLUSTER))
  
  return(list("exp"=exp, "df"=df, "cls"=clusters_list))
}

obj <- sync_metadata(x, skin_clinical)
exp <- obj$exp # exposure matrix in long format (removed clusters with less than 20 samples)
df <- obj$df # clinical data (samples from clinical data are mapped with samples from samples from bascule object)
df <- df %>% select(-AGE) %>% rename(AGE=AGE_GROUP) # remove age column from metadata
```


### Exploratory Analysis (exposures and de novo signatures)

Analysing 259 samples, two clusters (samples belonging to clusters with less than or equal 20 members eliminated) and 10 mutational signatures, include eight established signatures SBS1, SBS3, SBS5, SBS7a, SBS7c, SBS7d, SBS18 and SBS137 from COSMIC and Degasperi study, alongside two de novo signatures SBSD4 and SBSD8 detected using the BASCULE framework.
Cluster G1, comprising a smaller subset of samples (n = 37), exhibits a heterogeneous mutational profile, predominantly enriched in age-related signatures SBS1 and SBS5, along with SBS3 and SBS18 which has been associated with homologous recombination deficiency and damage by reactive oxygen species. This distinct mutational composition suggests an etiology that diverges from the canonical ultraviolet (UV) damage pathway. Additionally, SBSD8 contributes to the mutational landscape of cluster G1, with some samples displaying elevated proportions of these signatures. In contrast, cluster G11, encompassing the majority of samples (n = 190), is primarily characterized by the SBS7 family (SBS7a, SBS7c and SBS7d), strongly indicative of UV-induced mutagenesis, aligning with the well-established role of UV exposure in the pathogenesis of skin cancers such as melanoma. Notably, signature SBS137 (from Degasperi catalogue), currently with unknown aetiology, is present, and the contribution of the de novo signature SBSD4 is more pronounced in G11 than in G1. The presence of novel signatures SBSD4 and SBSD8 identified via BASCULE suggests the existence of previously uncharacterised mutagenic processes within this cohort.

```{r fig.height=6, fig.width=12, message=FALSE, warning=FALSE}
plot_exposures(x, clusters = c("G1", "G11"))
```



```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
#plot_signatures(x)
plot_signatures(x, signames = get_denovo_signames(x))
```


### Kaplan-Meier Analysis
In a study of n=259 skin cancer samples, we identified ten mutational signatures within two clusters, including the known established signatures like SBS1 (clock-like), SBS5 (clock-like) and SBS7 family (UV-light). Kaplan-Meier survival analysis revealed distinct survival trends between the clusters (log-rank p = 0.033). Cluster G11, which was enriched for SBS7 signatures, exhibited better survival compared to G1.


```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
# create survival object
fit <- survfit(Surv(time = df$OS_YEARS, event = df$OS_STATUS) ~ CLUSTER, data = df)

ggsurvplot(
  fit = fit, 
  data = df, 
  size = 1, 
  pval = TRUE, # show log-rank test p-value
  #conf.int = TRUE, # show confidence intervals
  risk.table = TRUE, # add risk table below plot
  risk.table.col = 'strata', 
  legend.labs = c("G1", "G11"), 
  title = "Kaplan-Meier Survival Curves", 
  xlab = "Time (Years)", 
  ylab = "Survival Probability", 
  legend.title = "Cluster", 
  #palette = "jco", #c('orange', '#2e9fdf') # color scheme
  #censor.shape = "|", cencor.size = 4, 
  pval.size = 3, 
  fontsize = 4, 
  ggtheme = theme_bw()
)
```

### Cox proportional hazards analysis
In multivariate Cox regression, with G1 as reference due to its lowest survival outcomes, cluster G11 showed a trend toward a protective effect (n = 190, hazard ratio = 0.63, 95% CI: 0.39-1.02, p = 0.059). This finding however was borderline significant, possibly because of the effect of age which significantly influenced the hazard: younger patients experienced a 44% reduction in risk relative to older individuals (n = 114, hazard ratio = 0.56, 95% CI: 0.40-0.80, p = 0.001), whereas gender did not significantly affect the outcome.

```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}

format <- forest_model_format_options(
    colour = "black",
    color = NULL,
    shape = 15,
    text_size = 5,
    point_size = 6,
    banded = TRUE
  )
cox_model <- coxph(Surv(time = OS_YEARS, event = OS_STATUS) ~ GENDER + CLUSTER + AGE, data = df)
forest_model(cox_model, format_options = format, factor_separate_line = T, recalculate_width = F, recalculate_height = T) + 
  labs(title = "Cox proportional hazards model (forest plot)", tag = "C") + 
  patchwork::plot_layout(axis_titles="collect", axes="collect")# + 
  #theme_text & theme_legend & theme(plot.caption=element_text(hjust=.5))
```

