---
title: "Inference"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(bascule)
```


```{r setup, eval=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
reticulate::install_python(version="3.9.16")
reticulate::py_install(packages="pybascule", 
                       pip=TRUE,
                       python_version="3.9.16")
py = reticulate::import("pybascule")
```

# Load datasets

We can load the data `example_dataset`, a bascule object containing the true signatures and exposures used to generate the mutation counts matrix. In the object, data for SBS and DBS is reported.
```{r}
data("synthetic_data")
```

We can extract the mutation count matrix from the object using the `get_input` function. With `reconstructed=FALSE` we are obtaining the observed counts, and not the reconstructed ones computed as the matrix multiplication of exposures and signatures.

```{r}
counts = synthetic_data$counts
head(counts[["SBS"]][, 1:10])
head(counts[["DBS"]][, 1:10])
```

We use as reference the COSMIC catalogue for SBS and DBS.

```{r}
reference_cat = list("SBS"=COSMIC_sbs_filt, "DBS"=COSMIC_dbs)
```

In the example dataset, the true number of signatures is five for both SBS and DBS, thus we can provide as list of K to test values from 0 to 7.

```{r}
K_list = 0:7
```

# Fit the model

Now, we can fit the model. Let's first fit the NMF to perform signatures deconvolution. 
```{r eval=FALSE}
x = fit(counts=counts, k_list=K_list, n_steps=3000,
        reference_cat=reference_cat,
        keep_sigs=c("SBS1","SBS5"), # force fixed signatures
        store_fits=TRUE, py=py)
```

```{r}
x = synthetic_data$x
```

# Visualize the inference scores

You can inspect the model selection procedure.
```{r}
plot_scores(x)
```

```{r}
plot_gradient_norms(x)
```


# Visualize the inferred parameters

You can visualize the inferred signatures. Here, we notice that from `r nrow(COSMIC_sbs_filt)` and `r nrow(COSMIC_dbs)` SBS and DBS signatures, the model only selects `r length(get_fixed_signames(x)$SBS)` and  `r length(get_fixed_signames(x)$DBS)` signatures from the catalogue. Moreover, it also infers a denovo signature from the SBS counts.

```{r}
plot_signatures(x)
```


# Post fit heuristics and clustering

We can notice from the signatures plots a clear similarity between certain signatures. 
We can compute a linear combination on de novo signatures to remove those similar to reference ones.
On the refined set of signatures we can run the model to perform clustering of samples based on exposures.

```{r eval=FALSE}
x_refined = refine_denovo_signatures(x)
x_refined_cluster = fit_clustering(x_refined, cluster=3)
```
```{r}
x_refined_cluster = synthetic_data$x_refined_cluster
```


# Visualise clustering results 


```{r}
plot_signatures(x_refined_cluster)
```


```{r}
plot_centroids(x_refined_cluster)
```

```{r}
plot_exposures(x_refined_cluster)
```


